\LARGE \bf GPU Acceleration for Fixed Complexity Sphere Decoder in Large MIMO Uplink Systems

### 2015-02-04 FCSD-CUDA project/The final draft of CUDA paper/conf_draft/conference draft.tex
Revise after acceptance version 2

% draft mode.
%\documentclass[a4paper, 10pt, conference]{ieeeconf}
\documentclass[letterpaper, 10pt, conference]{ieeeconf}
%\documentclass[12pt,a4paper,final]{article}
%\IEEEoverridecommandlockouts
%\overrideIEEEmargins
%\documentclass[conference]{IEEEtrans}
% Add the compsoc option for Computer Society conferences.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,ft.tex

% manually specify the path to it like:
\usepackage{amssymb}
%\usepackage{amsthm}
%\usepackage{caption}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{epstopdf}
%\usepackage{graphics} % for pdf, bitmapped graphics files
%\usepackage{epsfig} % for postscript graphics files
\usepackage{mathptmx} % assumes new font selection scheme installed
%\usepackage{times} % assumes new font selection scheme installed
%\usepackage{amsmath} % assumes amsmath package installed
\usepackage{amssymb}  % assumes amsmath package installed
\usepackage{fancyhdr}
% Some very useful LaTeX packages include:
% (uncomment the ones you want to load)

@@

\begin{document}
% make the title area
%\begin{spacing}{2.5}
%\date{}
\maketitle
\thispagestyle{fancy} % IEEE模板在\maketitle后会自动声明\thispagestyle{plain}，  
                      % 导致第一页什么都没有。所以得把plain更改为fancy  
\lhead{} % 页眉左，需要东西的话就在{}内添加  
\chead{} % 页眉中  
\rhead{\thepage} % 页眉右  
\lfoot{} % 页眉左  
\cfoot{} % 页眉中  
\rfoot{} %页眉右，\thepage 表示当前页码  
\renewcommand{\headrulewidth}{0pt} %改为0pt即可去掉页眉下面的横线  
\renewcommand{\footrulewidth}{0pt} %改为0pt即可去掉页脚上面的横线  
\pagestyle{fancy}  
\rhead{\thepage}
%\thispagestyle{empty}
%\pagestyle{empty}
% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
Large MIMO constitutes a principal technology for next generation wireless communication systems. Detection algorithms for large MIMO schemes tend to require long simulation times on conventional computer systems. In order to improve simulation efficiency of decoders in large MIMO systems, we propose to use General Purpose Graphic Processing Units (GPGPU). This paper considers the implementation of a Fixed Complexity Sphere Decoder (FCSD) for large scale MIMO systems using the framework of GPGPU-Computed Unified Device Architecture (CUDA). The proposed CUDA-FCSD can achieve considerable speedups over conventional CPU implementation while providing the same Bit Error Rate (BER) performance. Furthermore the proposed algorithm demonstrates the great potential of GPGPU to accelerate the simulation of large MIMO systems.

@@

General Purpose Graphic Processing Units (GPGPU) computing is a common trend in industry and research in areas that need massive computations, such as computer vision\cite{fung2008using}, computer games\cite{blewitt2013applicability}, signal processing\cite{van2011accelerating}\cite{6671435}, and finance\cite{grauer2013accelerating}. Compared to a CPU, a GPU can naturally cope with computational intensive tasks, since most of the hardware resources are allocated for data processing and parallel computation, rather than caching and flow control\cite{cook2013cuda}. In recent years GPU computing has provided significant advantages in data processing speeds over CPU computations\cite{nvidia2008programming}. The Computed unified Device Architecture (CUDA)\cite{cook2013cuda}\cite{nvidia2008programming} is a programming model for GPGPU that was proposed by NIVIDIA and has become the dominant framework for GPGPU programming. It is based on popular programming languages (C, C++, Java, Python, Fortran, Direct Compute), such that the application developers in different areas do not need to learn a new programming language or get to know the hardware details. CUDA employs the Single Instruction Multiple Threads (SIMT) programming model, enables a GPU to implement parallel data processing, and has become a parallel computing platform for different applications.

  A FCSD implementation based on GPU was presented in \cite{hongyuan2010fixed}, however this implementation only considers executing thousands of decoders in parallel rather than exploiting the parallel potential of FCSD itself. In this paper, a GPU based implementation of Fixed Complexity Sphere Decoder (FCSD) is proposed, which is referred to as CUDA-FCSD. A heterogeneous programming solution for FCSD preprocessing and integrated blocked parallel path searching and decision making are presented. Using this implementation the speedup as well as Bit Error Rate (BER) performance of different sizes of MIMO systems and signal constellations are analyzed. The proposed CUDA-FCSD implementation demonstrates a great potential for large MIMO systems.

@@

Threads are scheduled, managed, created and executed in a group of 32, called warp. The programs that execute on the device side are defined as \textit{kernel} functions. When a \textit{kernel} function is invoked, the thread blocks are partitioned into warps. The threads in one warp execute the same instruction simultaneously. The different warps are scheduled by a warp scheduler. When one warp is waiting for data loading, the other warps begin to execute. Such an approach is called a Single Instruction Multiple Thread (SIMT) model. This programming architecture is akin to a Single Instruction Multiple Data (SIMD) scheme, however unlike SIMD vectoring machines, the SIMT can branch the behavior of each threads to achieve thread-level parallelism. The GPU uses a massive number of threads to hide the long latency of a global memory, instead of using a large cache as CPU. A CPU core is designed to minimize the latency for each thread at a time, whereas a GPU is designed to handle a large number of concurrent lightweight threads, in order to maximize throughput.
\subsection{Mircroarchitecture of GPU} 
We present the GPU-GeForce GTX 760 microarchitecture in Fig. \ref{figure2} as an example. The GPU works as a coprocessor system, consisting of Stream Multiprocessors (SMs), which perform multiply-add arithmetic operations. Furthermore, each SM has the special functional units (SFU) that can perform more complex operations such as trigonometric functions and reciprocal square root. There are five types of memories: registers, shared memory, global memory, constant memory and texture memory. Registers are allocated dynamically and privately to threads, and provide most rapid access speeds. The shared memory, also called on-chip memory, is located on the SMs and can be shared among the threads in one block. Shared memory has low latency and limited capacity. Global memory, which is off-chip has large latency and large capacity. The other two types of memories, constant memory and texture memory, are located off-chip, but cached. Therefore the access speed of these two type of memory space is much faster than that of global memory. The variables invoked by \textit{kernel} functions can be stored in either of the these five types of memories.

@@

In conclusion the FCSD symbol vector candidate $s^{F}_{i}=[s_{1},s_{2}\dots s_{N_{t}}]$ can be expressed as:
\begin{equation}
\begin{split}
&FE \quad stage:\\
&{s}_{i}^{F}\in    \mathbb{O}^{\rho},  \quad if\quad i=N_{t},N_{t}-1,\dots N_{t}-\rho+1\\
& SE\quad stage:\\
&{s}_{i}^{F}=\mathbb{Q}[\hat{s}_{i}+\sum_{j=i+1}^{N_{t}}\frac{r_{ij}}{r_{ii}}(\hat{s}_{j}-s_{j}^{F})], \quad if\quad \\
& i = N_{t}-\rho,N_{t}-\rho-1,\dots 2,1 
  \label{FCSD solution}
\end{split}
\end{equation}
where $\mathbb{Q}[\cdot]$ denotes the operation of signal constellation quantization. At the post processing stage, the FCSD compares the Euclidean distance of all the symbol vector candidates
\begin{equation}
E_{metric}=||\mathbf{R}(\mathbf{\hat{s}}-\mathbf{s}^{F})||^{2},
\end{equation} 
where $\mathbf{s}^{F}$ denotes the symbol vector candidate. The symbol vector candidate with minimum $E_{metric}$ is chosen as the solution. The total number of candidates is $M^{\rho}$.  
\begin{figure}[htb]
\centering
\includegraphics[width=0.5\textwidth, height=8cm]{FCSD_tree_searching.eps}
\caption{Tree searching of 4-QAM FCSD for 8$\times$8 MIMO system. The total number of branches is $4^{2}$}
\label{figure4}
\end{figure}
\subsection{CUDA-FCSD Preprocessing}
A FCSD channel ordering strategy is used in order to avoid error propagation in the serial path searching process. This channel ordering strategy is based on post processing signal to noise ratio\cite{wolniansky1998v}
\begin{equation}
\varphi_{m}=\frac{E_{s}}{\sigma^{2}(\mathbf{H}^{H}\mathbf{H})_{mm}^{-1}},  \label{ppsnr}
\end{equation}
where $\varphi_{m}$ denotes the post processing signal to noise ratio of the $m$th data stream, and $(\mathbf{H}^{H}\mathbf{H})_{mm}^{-1}$ denotes the diagonal elements of the inverse of $\mathbf{H}^{H}\mathbf{H}$. The post processing SNR depends only on $(\mathbf{H}^{H}\mathbf{H})_{mm}^{-1}$. From \cite{barbero2008fixing}, at the FE stage the "weakest" data stream, that is the data stream with the smallest $\varphi_{m}$, is detected first. At the SE stage, because of the single searching branch of the node, the performance is tightly related to the post processing SNR. Therefore at the SE stage the data stream with the largest post processing SNR is detected first in order to avoiding error propagation. In conclusion the FCSD ordering works iteratively  and obeys the following rule:
at the $j$th step ($j=1,2\dots N_{t}-1$), $s_{p}$ denotes the symbol to be detected, where 
\begin{equation}
p=\left\lbrace \begin{array}{c}
arg\max_{k}(\mathbf{H}_{j}^{H}\mathbf{H}_{j})_{kk}^{-1}\quad FE\quad stage\\
arg\min_{k}(\mathbf{H}_{j}^{H}\mathbf{H}_{j})_{kk}^{-1}\quad SE\quad stage   \label{the ordering}
\end{array}\right.  
\end{equation}  
and $\mathbf{H}_{j}\in \mathbb{C}^{N_{r}\times (N_{t}-j+1)}$ denotes the renewed channel matrix with the column corresponding to previous detected data streams removed from $\mathbf{H}_{j-1}$.

@@

Global memory access is implemented by a dynamic random access memory (DRAM) whose R/W is an extremely slow process. In a modern DRAM the R/W operations are performed by accessing a piece of consecutive memory location. As mentioned in section \ref{programming model}, the warp is the basic schedule unit of a device, and memory space is accessed warp by warp. Thus the most effective global memory access pattern is to ensure a warp of threads can access a consecutive memory space. For the GeForce GTX 760, the DRAM will access a consecutive memory space of 128 bytes at one time. In Fig. \ref{coalesce global memory} we present the global access pattern of $\mathit{s_{pm}}$ as an example to illustrate global memory access pattern.
\begin{figure}[htb]
\centering
\includegraphics[width=0.5\textwidth, height=8cm]{coalescing_global_memory.eps}
\caption{Global Memory Access Pattern of $\mathit{s_{pm}}$}
\label{coalesce global memory}
\end{figure}

@@

In Fig. \ref{BER curve} we present Monte Carlo simulation results for bit error rate (BER) performance, using the GPU implementation of different MIMO system configurations and constellation sizes. We consider uncoded complex MIMO systems with $M$QAM, where each antenna transmits different streams of binary data generated randomly and independently. The results have been obtained by using $100000$ channel realizations with at least $500$ symbol errors accumulated. The simulation results show that FCSD-CUDA provides the same BER performance as its CPU version\cite{barbero2008fixing}.
\begin{figure}[htb]
\centering
\includegraphics*[width=0.56\textwidth]{BER_curves.eps}
\caption{BER performance for MIMO systems with various modulation schemes using the GPU implementation}
\label{BER curve}
\end{figure}
\section{Conclusions}\label{conclusion}
This paper presents a GPU implementation of the fixed complexity sphere decoder (FCSD) for large MIMO systems. In order to exploit the computational capability of GPU in the simulation of large MIMO systems, the utilization of a heterogeneous programming model, accounting for resource limitations and memory configuration must be considered. The simulation results show that a considerable acceleration of the GPU implementation of FCSD can be reached for large MIMO systems and signal constellation sizes over CPU implementation, while maintaining the same BER performance. Since the GPU and CPU can work independently, the data processing power of GPU computing is an additional advantages provided by the traditional simulation platforms. This shows the potential of GPU computing to reduce the time associated with intensive simulations of large MIMO systems.

>>>>

% draft mode.
%\documentclass[a4paper, 10pt, conference]{ieeeconf}
\documentclass[letterpaper, 10pt, conference,twoside]{ieeeconf}
%\documentclass[12pt,a4paper,final]{article}
%\IEEEoverridecommandlockouts
%\overrideIEEEmargins
%\documentclass[conference]{IEEEtrans}
% Add the compsoc option for Computer Society conferences.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,ft.tex

% manually specify the path to it like:
\usepackage{amssymb}
%\usepackage{amsthm}
%\usepackage{caption}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{epstopdf}
%\usepackage{graphics} % for pdf, bitmapped graphics files
%\usepackage{epsfig} % for postscript graphics files
\usepackage{mathptmx} % assumes new font selection scheme installed
%\usepackage{times} % assumes new font selection scheme installed
%\usepackage{amsmath} % assumes amsmath package installed
\usepackage{amssymb}  % assumes amsmath package installed
\usepackage{fancyhdr}
\usepackage{lastpage}
% Some very useful LaTeX packages include:
% (uncomment the ones you want to load)

@@

\begin{document}
% make the title area
%\begin{spacing}{2.5}
%\date{}
\maketitle
\thispagestyle{fancy} % IEEE模板在\maketitle后会自动声明\thispagestyle{plain}，  
                      % 导致第一页什么都没有。所以得把plain更改为fancy
%\fancyhead{\thepage} % 奇数页右侧，偶数页左侧显示页码 
\lhead{} % 页眉左，需要东西的话就在{}内添加  
\chead{} % 页眉中  
\rhead{} % 页眉右  
\lfoot{} % 页眉左  
\cfoot{} % 页眉中  
\rfoot{} %页眉右，\thepage 表示当前页码  
\renewcommand{\headrulewidth}{0pt} %改为0pt即可去掉页眉下面的横线  
\renewcommand{\footrulewidth}{0pt} %改为0pt即可去掉页脚上面的横线  
\pagestyle{fancy}  
\fancyhead[RO,LE]{\thepage}  
%\rhead{\thepage}
%\thispagestyle{empty}
%\pagestyle{empty}
% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
Large MIMO constitutes a principal technology for next generation wireless communication systems. Detection algorithms for large MIMO schemes tend to require long simulation times on conventional computer systems. In order to improve simulation efficiency of decoders in large MIMO systems, we propose to use General Purpose Graphic Processing Units (GPGPU). This paper considers the implementation of a Fixed Complexity Sphere Decoder (FCSD) for large scale MIMO systems using the framework of GPGPU-Computed Unified Device Architecture (CUDA). Significant speedups were formed for large MIMO systems and constellation sizes over conventional CPU implementation while providing the same Bit Error Rate (BER) performance, hence the proposed algorithm provides great potential for GPGPU to accelerate the simulation of large MIMO systems.

@@

General Purpose Graphic Processing Units (GPGPU) computing is a common trend in industry and research in areas that need massive computations. Compared to a CPU, a GPU can naturally cope with computational intensive tasks, since most of the hardware resources are allocated for data processing and parallel computation, rather than caching and flow control\cite{cook2013cuda}. In recent years GPU computing has provided significant advantages in data processing speeds over CPU computations\cite{nvidia2008programming}. GPU computing has been applied to different research fields such as computer vision\cite{fung2008using}, computer games\cite{blewitt2013applicability}, signal processing\cite{van2011accelerating}\cite{6671435}, and finance\cite{grauer2013accelerating}. The Computed unified Device Architecture (CUDA)\cite{cook2013cuda}\cite{nvidia2008programming} is a programming model for GPGPU that was proposed by NIVIDIA and has become the dominant framework for GPGPU programming. It is based on popular programming languages (C, C++, Java, Python, Fortran, Direct Compute), such that the application developers in different areas do not need to learn a new programming language or get to know the hardware details. CUDA employs the Single Instruction Multiple Threads (SIMT) programming model, enables a GPU to implement parallel data processing, and has become a parallel computing platform for different applications.

 \cite{hongyuan2010fixed} presents an interesting implementation of Fixed Complexity Sphere Decoder (FCSD) based on GPU, however only parallel implementation of multiple decoders is considered rather than exploiting the attribute of parallelism of the algorithm itself. In this paper, a GPU based implementation of single FCSD is proposed, which is referred to as CUDA-FCSD. A heterogeneous programming solution for FCSD preprocessing and integrated blocked parallel path searching and decision making are presented. Using this implementation the speedup as well as Bit Error Rate (BER) performance of different sizes of MIMO systems and signal constellations are analyzed. The proposed CUDA-FCSD implementation demonstrates a great potential for large MIMO systems.

@@

Threads are scheduled, managed, created and executed in a group of 32, called warp. The programs that execute on the device side are defined as \textit{kernel} functions. When a \textit{kernel} function is invoked, the thread blocks are partitioned into warps. The threads in one warp execute the same instruction simultaneously. When one warp is waiting for data loading, the other warps begin to execute. Such an approach is called a Single Instruction Multiple Thread (SIMT) model, whereby the behavior of each thread can be branched to achieve thread level parallelism. The GPU uses a massive number of threads to hide the long latency of a global memory, instead of using a large cache as CPU. A CPU core is designed to minimize the latency for each thread at a time, whereas a GPU is designed to handle a large number of concurrent lightweight threads, in order to maximize throughput.
\subsection{Mircroarchitecture of GPU} 
We present the GPU-GeForce GTX 760 microarchitecture in Fig. \ref{figure2} as an example. The GPU works as a coprocessor system, consisting of Stream Multiprocessors (SMs), which perform multiply-add arithmetic operations. Furthermore, each SM has the special functional units (SFU) that can perform more complex operations such as trigonometric functions and reciprocal square root. There are five types of memories: registers, shared memory, global memory, constant memory and texture memory. Registers are allocated dynamically and privately to threads, and provide most rapid access speeds. The shared memory, also called on-chip memory, is located on the SMs and can be shared among the threads in one block. Shared memory has low latency and limited capacity. Global memory, which is off-chip has large latency and large capacity. The other two types of memories, constant memory and texture memory, are located off-chip, but cached. Therefore the access speed of these two type of memory space is much faster than that of global memory. The variables invoked by \textit{kernel} functions can be stored in either of the these five types of memories.

@@

In conclusion the FCSD symbol vector candidate $s^{F}_{i}=[s_{1},s_{2}\dots s_{N_{t}}]$ can be expressed as:
\begin{equation}
\begin{split}
&FE \quad stage:\\
&{s}_{i}^{F}\in    \mathbb{O}^{\rho},  \quad if\quad i=N_{t},N_{t}-1,\dots N_{t}-\rho+1\\
& SE\quad stage:\\
&{s}_{i}^{F}=\mathbb{Q}[\hat{s}_{i}+\sum_{j=i+1}^{N_{t}}\frac{r_{ij}}{r_{ii}}(\hat{s}_{j}-s_{j}^{F})], \quad if\quad \\
& i = N_{t}-\rho,N_{t}-\rho-1,\dots 2,1 
  \label{FCSD solution}
\end{split}
\end{equation}
where $\mathbb{Q}[\cdot]$ denotes the operation of signal constellation quantization. At the post processing stage, the FCSD compares the Euclidean distance of all the symbol vector candidates
\begin{equation}
E_{metric}=||\mathbf{R}(\mathbf{\hat{s}}-\mathbf{s}^{F})||^{2},
\end{equation} 
where $\mathbf{s}^{F}$ denotes the symbol vector candidate. The symbol vector candidate with minimum $E_{metric}$ is chosen as the solution. The total number of candidates is $M^{\rho}$.  
\begin{figure}[htb]
\centering
\includegraphics[width=0.5\textwidth, height=8cm]{FCSD_tree_searching.eps}
\caption{Tree searching of 4-QAM FCSD for 8$\times$8 MIMO system. The total number of branches is $4^{2}$}
\label{figure4}
\end{figure}
\subsection{CUDA-FCSD Preprocessing}
A FCSD channel ordering strategy is used in order to avoid error propagation in the serial path searching process. This channel ordering strategy is based on post processing signal to noise ratio\cite{wolniansky1998v}
\begin{equation}
\varphi_{m}=\frac{E_{s}}{\sigma^{2}(\mathbf{H}^{H}\mathbf{H})_{mm}^{-1}},  \label{ppsnr}
\end{equation}
where $\varphi_{m}$ denotes the post processing signal to noise ratio of the $m$th data stream, and $(\mathbf{H}^{H}\mathbf{H})_{mm}^{-1}$ denotes the diagonal elements of the inverse of $\mathbf{H}^{H}\mathbf{H}$. The post processing SNR depends only on $(\mathbf{H}^{H}\mathbf{H})_{mm}^{-1}$. From \cite{barbero2008fixing}, at FE stage the data stream with minimum post processing SNR is detected first, while at the SE stage the data stream with maximum post processing SNR is detected first. In conclusion the FCSD ordering works iteratively and obeys the following rule:
at the $j$th step ($j=1,2\dots N_{t}-1$), $s_{p}$ denotes the symbol to be detected, where 
\begin{equation}
p=\left\lbrace \begin{array}{c}
arg\max_{k}(\mathbf{H}_{j}^{H}\mathbf{H}_{j})_{kk}^{-1}\quad FE\quad stage\\
arg\min_{k}(\mathbf{H}_{j}^{H}\mathbf{H}_{j})_{kk}^{-1}\quad SE\quad stage   \label{the ordering}
\end{array}\right.  
\end{equation}  
and $\mathbf{H}_{j}\in \mathbb{C}^{N_{r}\times (N_{t}-j+1)}$ denotes the renewed channel matrix with the column corresponding to previous detected data streams removed from $\mathbf{H}_{j-1}$.

@@

Global memory access is implemented by a dynamic random access memory (DRAM) whose R/W is an extremely slow process. In a modern DRAM the R/W operations are performed by accessing a piece of consecutive memory location. As mentioned in section \ref{programming model}, the warp is the basic schedule unit of a device, and memory space is accessed warp by warp. Thus the most effective global memory access pattern is to ensure a warp of threads can access a consecutive memory space. For the GeForce GTX 760, the DRAM will access a consecutive memory space of 128 bytes at one time. In Fig. \ref{coalesce global memory} we present the global access pattern of $\mathit{s_{pm}}$ as an example to illustrate global memory access pattern.
\begin{figure}[htb]
\centering
\includegraphics[width=0.5\textwidth, height=6cm]{coalescing_global_memory.eps}
\caption{Global Memory Access Pattern of $\mathit{s_{pm}}$}
\label{coalesce global memory}
\end{figure}

@@

In Fig. \ref{BER curve} we present Monte Carlo simulation results for bit error rate (BER) performance, using the GPU implementation of different MIMO system configurations and constellation sizes. We consider uncoded complex MIMO systems with $M$ QAM, where each antenna transmits different streams of binary data generated randomly and independently. The results have been obtained by using $10^{5}$ channel realizations with at least $500$ symbol errors accumulated. The simulation results show that FCSD-CUDA provides the same BER performance as its CPU version, due to the space limitation we do not present the results of CPU implementation, \cite{barbero2008fixing} provides some of the referable results.
\begin{figure}[htb]
\centering
\includegraphics*[width=0.5\textwidth, height=10cm]{BER_curves.eps}
\caption{BER performance for MIMO systems with various modulation schemes using the GPU implementation}
\label{BER curve}
\end{figure}
\section{Conclusions}\label{conclusion}
This paper presents a GPU implementation of the fixed complexity sphere decoder (FCSD) for large MIMO systems. In order to exploit the computational capability of GPU in the simulation of large MIMO systems, the utilization of a heterogeneous programming model, accounting for resource limitations and memory configuration must be considered. The simulation results show that significant accelerations of the GPU implementation of FCSD can be reached for large MIMO systems and signal constellation sizes over CPU implementation, while maintaining the same BER performance. Since the GPU and CPU can work independently, the data processing power of GPU computing is an additional advantages provided by the traditional simulation platforms. This shows the potential of GPU computing to reduce the time associated with intensive simulations of large MIMO systems.
